\section{Validation Results}

Validation tests compare computed visibilities against analytical predictions.

\subsection{Phase and Amplitude Validation}

For a point source at angle $\theta$ observed by antennas separated by baseline $d$, the expected visibility phase is $\phi = 2\pi f d \sin\theta / c$. Testing across varied configurations (baselines 10–20m, angles 0–60°, frequencies 710–1420 MHz) yielded phase errors below $0.2^\circ$, well within requirements. Errors arise from numerical precision and FFT discretization.

Closure phases for four-antenna triangles measured $< 0.03^\circ$, confirming phase coherence across all baselines independent of calibration.

Amplitude validation using calibrated sources showed measured SNR matching predictions within 3\%. SNR scales as $\sqrt{\Delta t}$ as expected, confirming proper integration.

\subsection{Multi-Source and Noise Performance}

Two-source testing (sources at 0° and 30° with fluxes 1.0 and 0.5 Jy) verified correct superposition. Measured visibility amplitudes and phases matched vector sum predictions within 1\%, confirming linearity.

Noise performance testing across SNR 0.1–100 showed: strong signals (SNR $> 10$) maintain phase precision $< 6^\circ$; marginal detections (SNR $\approx$ 3) show $19^\circ$ uncertainty; SNR $< 1$ becomes undetectable as expected. Phase uncertainty follows $\sigma_\phi \approx 1/\mathrm{SNR}$ (radians), matching theory.

\subsection{Spectral and Temporal Validation}

Narrowband source testing confirmed spectral localization: signal appears exclusively in the correct channel with neighbouring channels containing only noise, demonstrating accurate channelization without excessive leakage.

Integration time validation (0.1–100 s) showed noise scaling as $1/\sqrt{\Delta t}$ within 2\% of predictions, confirming proper accumulation. Mean visibility phases remained stable ($< 0.5^\circ$ standard deviation) across integration times.

Across all validation tests—single/multiple sources, various SNR, frequency channels, integration times—computed visibilities match analytical predictions within measurement uncertainty. Phase errors $< 0.2^\circ$ for strong signals, amplitudes within 1\%, closure phases confirming consistency, and noise following theoretical expectations establish that the implementation correctly performs FX correlation and produces scientifically valid outputs.
