\documentclass[a4paper,oneside,11pt]{book}
\usepackage{helvet}
\usepackage[hidelinks]{hyperref}
\usepackage[T1]{fontenc}
\renewcommand\familydefault{\sfdefault}
\usepackage{NWUStyle}
\usepackage{longtable}

% -----------------------------------------

\begin{document}

% -----------------------------------------

\Title{Four Element Correlator for the Nooitgedacht Radio Telescope}
\Initials{F. J.}
\FirstName{Francois}
\Surname{Meiring}
\StudentNumber{38276909}
\ORCID{0009-0004-7605-0618}
\MScorPhD{Magister Scientiae}
\Field{Computer Science}
\GradDate{April, 2027}
\Supervisor{Prof. Günther Drevin}
\newcommand{\cosupervisor}{Mr. J. Heystek Grobler}
\newcommand{\assistantsupervisor}{}

\newcommand{\keywords}[1]{\vspace{1em}\noindent\textbf{Keywords: }#1}

\date{\today}

% -----------------------------------------

\MakeTitle

% -----------------------------------------

\begin{Acknowledgements}{}
    I would like to thank the North-West University for the provision of resources and facilities to carry out this research. The staff and researchers at the Nooitgedacht Radio Telescope facility are thanked for their cooperation and support during this project.
    
    I wish to thank my supervisors, Prof. Günther Drevin and Mr. J. Heystek Grobler, for their guidance, expertise, and continued support throughout the process of realising this digital correlator system. They have made constructive suggestions and comments that have influenced both the technical implementation and the research direction.
    
    I am grateful to the radio astronomy community, and specifically those researchers whose work in software correlators and Python-based implementations has laid the groundwork for this project. Their papers and open-source contributions have enabled this work.
    
    I also appreciate the student and technical colleagues who assisted on various aspects of the project, from hardware integration to software testing. Their hands-on involvement and collaborative spirit have enhanced the value of this work.
    
    Finally, I would also appreciate my friends and relatives for their encouragement and support during my study without wavering. Their patience and tolerance have been priceless to the completion of this thesis.
\end{Acknowledgements}

% -----------------------------------------

\input{sections/Abstract.tex}

\keywords{Telescope, Simulator, Physics, Science, Astronomy}

% -----------------------------------------

\tableofcontents
\listoffigures

% \MakeTOCandLOFandLOT % Create table of contents, list of figures, and list of tables

\begin{TableOfAbbrev}
    \begin{longtable}{ll}
        \textbf{ACS} & Antenna Control Subsystem \\
        \textbf{ADC} & Analogue-to-Digital Converter \\
        \textbf{AI} & Artificial Intelligence \\
        \textbf{ALMA} & Atacama Large Millimeter/submillimeter Array \\
        \textbf{ASIC} & Application-Specific Integrated Circuit \\
        \textbf{CAM} & Control and Monitoring \\
        \textbf{CASPER} & Collaboration for Astronomy Signal Processing and Electronics Research \\
        \textbf{CHIME} & Canadian Hydrogen Intensity Mapping Experiment \\
        \textbf{CPU} & Central Processing Unit \\
        \textbf{CuPy} & CUDA Python (NumPy-compatible GPU array library) \\
        \textbf{DiFX} & Distributed FX Software Correlator \\
        \textbf{DOF} & Degree of Freedom \\
        \textbf{FFT} & Fast Fourier Transform \\
        \textbf{FIR} & Finite Impulse Response \\
        \textbf{FPGA} & Field-Programmable Gate Array \\
        \textbf{FRB} & Fast Radio Burst \\
        \textbf{FX} & Fourier Transform followed by Cross-multiplication Architecture \\
        \textbf{FXF} & Filter-FX Architecture \\
        \textbf{GPS} & Global Positioning System \\
        \textbf{GPU} & Graphics Processing Unit \\
        \textbf{HDF5} & Hierarchical Data Format version 5 \\
        \textbf{I/O} & Input/Output \\
        \textbf{JIT} & Just-In-Time Compilation \\
        \textbf{KATCP} & Karoo Array Telescope Communication Protocol \\
        \textbf{LNA} & Low-Noise Amplifier \\
        \textbf{LQG} & Linear Quadratic Gaussian \\
        \textbf{MPC} & Model Predictive Control \\
        \textbf{ML} & Machine Learning \\
        \textbf{NTP} & Network Time Protocol \\
        \textbf{NumPy} & Numerical Python \\
        \textbf{PFB} & Polyphase Filter Bank \\
        \textbf{PID} & Proportional-Integral-Derivative Controller \\
        \textbf{PPS} & Pulse Per Second \\
        \textbf{PyCUDA} & Python CUDA Interface \\
        \textbf{RA} & Right Ascension \\
        \textbf{RTOS} & Radio Telescope Orientation Software \\
        \textbf{RFI} & Radio Frequency Interference \\
        \textbf{SDR} & Software-Defined Radio \\
        \textbf{SKA} & Square Kilometre Array \\
        \textbf{SKARAB} & SKA Reconfigurable Application Board \\
        \textbf{SNAP} & Smart Network ADC Processor \\
        \textbf{UDP} & User Datagram Protocol \\
        \textbf{VDIF} & VLBI Data Interchange Format \\
        \textbf{VLA} & Very Large Array \\
        \textbf{VLBA} & Very Long Baseline Array \\
        \textbf{xGPU} & GPU-Based Cross-Correlation Library \\
    \end{longtable}
\end{TableOfAbbrev}

% -----------------------------------------

\pagestyle{headings}
\setcounter{page}{1}
\pagenumbering{arabic}

% -----------------------------------------

\input{sections/Introduction.tex}
\input{sections/Research Methodology & Paradigm}
\input{sections/Literature Review/Main.tex}
\input{sections/Conclusion.tex}

% -----------------------------------------

\bibliography{Bibliography}
\appendix

% -----------------------------------------

\end{document}